---
title: "Example #1: Calibration Protocal - Time-Averaged"
author: Julia L. Blanchard
date: July 2020
place: Hobart, Australia
output:
  pdf_document: default
  html_document: default
---

# Calibrating a multi-species model to time-averaged species' catches

In this example we will explore how we can learn about models by fitting size spectrum ecological models to data using the "mizer" R package. 

Recall, there are three different kinds of size spectrum models in mizer, of increasing complexity: 
1) community model: purely size-based and representative of a single but  "average" species across the whole community 
2) trait-based model, which disaggregates the size spectrum into differentgroups with different life-histories, through differences in each "species" asymptotic which determines
other life-history parameters such as the size at maturity (Hartvig et al. 2011, Andersen & Pedersen, 2010)
3) multispecies model - which has the same equations and parameters as the trait-based model but is parameterised to represent multiple species in a real system, where each species can have many differing species-specific traits (Blanchard et al. 2014). 

Here we focus on multispecies size spectrum models. In practice, these models have been paramterised in a few different ways depending on data availability for a system or reserach questions.

Some studies have focused on  many species-specific values, for example where each species have different values of life-history, size-selective feeding trait parameters (e.g. \beta and \sigma), and details of species interactions (Blanchard et al. 2014, Reum et al. 208) to better capture the dynamics of marine food webs. 

Others, such as Jacobsen et al. (2014,2016), have represented variation in only a couple of the most important life history parametersfor each species - asymptotic size (which links to other parameters such as maturation size and ) and recruitment parameters (Rmax, eRepro) to broadly capture fished communities or carry out across ecosystem comparisons. 

Once you have paramterised the multispecies model for your system, you may find that species do not coexist or the biomass or catches are very different from your observations. After the model is parameterised and asssessed for basic principles and coexistence (see section x - Gustav?), further calibration to observational data is used to ensure the relative abundance of each species reflects the system, atleast during a stable period, which is time-averaged. 

The background resource parameters and the recruitment parameters, \Rmax (maximum recruitment) and \erepro (reproductive efficiency) greatly affect the biomasses of species in your system. 

The recruitment parameters are highly uncertain and capture density dependent processes in the model that limit the number of offspring that successfully recruit to the smallest size class for each species. In the default mizer package these paramters are used to implement an emergent Beverton-Holt type stock recruitment relationship. 

As a starting point, we will estimate these parameters as a means of fitting the modelled species catches to the observed catches. This could similarly be carried out with biomasses. Other model detailed approaches also exist, see the main paper, but this approach has been used to get models in the right "ball-park", which can them be further evaluated using diagnostics (example X) and fitted to time series data (example XX).


### A Simple Protocol for Multispecies Model Calibration

We will adapt the "recipe" for calibration in Jacobsen et al 2014 (see supp. mat.) and Blanchard et al (2014), into the following steps:


0. Run the model with the chosen species-specific parameters. This will relate some of the missing parameters to \Winf ( h and gamma - explain through simple example of how the model works?). \Rmax (see example that explains Rmax?) could also be automatically calculated based on equilbrium assumptions (Andersen et al. 2016) but by default it is "Inf", which means there is no density dependence asscoaited with spawner-recruit dynamics.

1. Obtain the time-averaged data (e.g. catches or biomasses for each species) and the time-averaged fishing mortalty inputs (e.g. from stock assessments). Typically this should be over a stable part of the time series for your system.

2. Start with the chosen parameters for kappa and lambda of the resource spectrum that are obtained from the literature regarding the community size spectrum. These can be very uncertain and sometimes are not available. Calibrate the carrying capacity of the background resource spectrum, \kappa, by examining the feeding level, biomass through time, and overall size spectrum. 

3. Calibrate the maximum recruitment, \Rmax, which will affect the relative biomass of each species (and, combined with the fishing parameters, the catches) by minimising the error between observed and estimated catches (again or biomasses).

4. Check that the physiological recruitment, \RDI, is much higher than the realised recruitment, \RDD. This can be done using the getRDD and getRDI functions and calculating the ratio which should be around 100  for a species with \Winf = 1500 g, but varies with asymptotic size and fishing mortality (Andersen 2019). High RDI/RDD ratio indicates the carrying capacity is controlling the population rather than predation or competition. Larger species often require more of this density dependent control than smaller ones. If RDI/RDD is too high, the efficiency of reproduction (erepro) can be lowered to ensure species do not outcompete others or or over-resilient to fishing. Lowering erepro biologically means higher egg mortality rate or wasteful energy invested into gonads. If RDI/RDD = 1 the species is in the linear part of the stock recruitment relationship (no spawner-recruit density dependence).

5. Verify the model after the above step by comparing the model with: species biomass or abundance distrubutions, feeding level, naturality mortality, growth, vulnerablity to fishing (fmsy) and catch, diet composition. Many handy functions for plotting these are available here: https://sizespectrum.org/mizer/reference/index.html

6. The final verification step is to force the model with time-varying fishing mortality to assess whether changes in time series in biomassess and catches capture observed trends. The model will not cpature all of the fluctuations from environmental processes (unless some of these are included), but should match the magnitude and general trend in the data.

#### Step 0. Run the model with the chosen species-specific parameters. This will relate some of the missing paramsters to \Winf. \Rmax will also be automatically calculated based on equilbrium assumptions (Andersen et al. 2016).


Let's read in the North Sea model parameters, stored in mizer.

```{r}
# Initial set up

# only need to do the installations once.
#install.packages("remotes")
library(remotes)
#remotes::install_github("sizespectrum/mizerExperimental")
library(mizerExperimental)

library(tidyverse)
library(ggplot2)
library(plotly)

# Let's use your params instead of  the North Sea model parameters, but change the parameters and assumptions so that it's essentially a different uncalibrated model, where species are not coexisting

sparams <- read.csv("data/df_param_changes Naj_PP_new.csv")
# sometimes worth checking what the model params would do if it was a trait-based version - eg life-history difference only
# sparams$beta <-100
# sparams$sigma <-1.5
# sparams$r_max=Inf
# int <- inter
# int<-1
inter<- read.csv("data/theta1_new.csv")[,-1]
params <- newMultispeciesParams(sparams, inter,kappa = 1e11,max_w=1e6)

# note the volume of this model is set to the reflect the entire volume of the North Sea - hence the very large kappa value. This is system specific and you need to check/think about what units you are using here.

#  You can add other params just for info
#  param$Volumecubicmetres=5.5e13    #unit of volume. Here total volume of North sea is used (Andersen & Ursin 1977)

# have a look at species parameters that have been calculated
species_params(params)

# have a look at fishing params
gear_params(params)

#lets' change the plotting colours
# library(viridis)
# params@linecolour[1:12] <-plasma(12)
# params@linecolour["Resource"] <-"seagreen3"

# run with fishing
sim <- project(params, t_max = 100, effort = 1,dt=0.1)
# plot
plot(sim)

# could  take a peak at other plots here to ensure species params are all ok:


plotlyGrowthCurves(sim,percentage = T)  # check these
plotlyFeedingLevel(sim)  # low feeding levels at large sizes for some groups
plotGrowthCurves(sim,"Trichiurus_lepturus") # check each species - not reacching maturation sizes - think about the params that  control feeding, though looking at next plot they seem to accumulate in a middle range size class

plotlySpectra(sim,power=1,total = T)
plotDiet(params,"Trichiurus_lepturus") ## perhaps feeding too much on background resource, though can see a shift at larger sizes, possible low on food when it shifts to other fish?? although depends whethe your consider it ot be "other food" too, such as benthos (as in North Sea model.



```

You get coexistence - is this becuase you have already calibrated Rmax using an erepro of 1?


#### Step 1. Obtain the time-averaged data (e.g. catches or biomasses for each species) and the time-averaged fishing mortalty inputs (e.g. from stock assessments). Typically this should be over a stable part of the time series for your system.


```{r}

#read in time-averaged  catches  
#cdat<-read.csv("data/time-averaged-catches.csv") ### check units: tonnes

# looks like your yields are stored in params object
params@species_params$yield


# the fishing moratlty rates are already stored in the param object as
params@species_params$catchability

# # let's start again and replace with the initial pre-calibration "guessed" Rmax and Kappa ---ignoring this for now, but you may want to look into these as well

# params@resource_params$kappa = 1e11
# # penalise the large species with higher density dependence
# params@species_params$R_max <- params@resource_params$kappa*params@species_params$w_inf^-1
# # and reduce erepro
# params@species_params$erepro[]<- 1e-3
# 
# params <- setParams(params)
# # run with fishing
# sim <- project(params, t_max = 100, effort =1)
# plot(sim)


## could compare with Gustav's projectToSteady but not sure i fully undertsand that function
# steadyparams <- projectToSteady(params, t_max = 100, return_sim = F)
# sim2 <- project(steadyparams, t_max = 100, effort =1)
# plot(sim2)
# plotlySpectra(sim2,power=1,total = T)

# here check that  community size spectrum (black line) is continuation of the background - looks to be

```

Species are coexisting. This is in part because we applied a stronger  Rmax effect for larger species. You can play with the above parameters but but it would take a lot of trial an error to achieve the right combination to get the biomass or catches similar to the observations.

#### Step 2. Calibrate the carrying capacity of the background resource spectrum, \kappa at steady state 

We could explore the effects further using Rshiny app, where we also have a plot of the biomass or catch data. First let's look at the basic diagnostics and tune kappa and erepro to make sure the feeding levels are high enough for each species and that biomasses coexist. 

```{r}
# Optional
# adjust Rmax and/or reproductive efficiency to examine whether improved steady state is achievable that way
library(shiny)
runApp("shiny-equilibrium")
# is there a way to save the final chosen values?


## see how simple changes in kappa and erepro params affect: coexistence,  feeding level > 0.5, predation mortality curves
## 

```

This improves matters a little, but we need to make some species-specific adjustments.

The shiny app helps with understanding the model but it is tricky to arrive at the best fit especially if we want to change several species parameter combinations at a time.

Let's choose some values that enable the most species to coexist as a starting point for optimisation. Note we won't vary erepro at the same time as Rmax (they depend on each other). However we will use the value of erepro selected form the shiny app.

#### Step 3. Calibrate the maximum recruitment, \Rmax, which will affect the relative biomass of each species (and, combined with the fishing parameters, the catches) by minimising the error between observed and estimated catches or biomasses. We could also include kappa in our estimation here (as in Blanchard et al 2104 & Spence et al 2016) but instead we will use the value that seemed OK in terms of feeding levels in the shiny app, roughly log10(11.5). Same goes for erepro, a value of 1e-2 seemed ok.


This might take AWHILE. Go watch some Netflix.

```{r}
# # change kappa and erepro based on shiny exploration, set up initial values based on "close to" equilibrium values from above sim
#   steadyparams@species_params$erepro[] <-1e-0
#   params@initial_effort<-1
#   params@initial_n<-sim@n[100,,]
#   params@initial_n_pp<-sim@n_pp[100,]
#   params <- setParams(params,kappa=10^(11))

# define the initial parameters to send to optimisation code below

# we need 12 Rmaxs, log10 scale
vary <- log10(params@species_params$R_max)
nspp <-dim(params@species_params)[1]
#vary<-runif(10,3,12) # or use completley made up values, same for each species test for effects of initial values

## set up the enviornment to keep the current state of the simulations 
state <- new.env(parent = emptyenv())
state$params <-  steadyparams

## the following getError function combines the steps of the optimisastion above - this time with the multispecies model and output the predicted size spectrum

## update below with project_steady and saving the state from each iteration

### I changed this to use your yield values

getError <- function(vary,params=params,dat=sparams$yield,env=state,data_type="catch", tol = 0.1,timetorun=10) {
  
  #env$params@species_params$R_max[]<-10^vary[1:nspp]
  params@species_params$R_max[]<-10^vary[1:nspp]
  
  params <- setParams(params)
  # run to steady state and update params
  # env$params<- projectToSteady(env$params, distance_func = distanceSSLogN,
  #                 tol = tol, t_max = 200,return_sim = F)
  # params<- projectToSteady(params, distance_func = distanceSSLogN,
  #                  tol = tol, t_max = 200,return_sim = F)
  # 
  # create sim object 
   
  sim <- project(params, effort = 1, t_max = timetorun) #Change t_max to determine how many years the model runs for
  
  # 
  # sim <- project(env$params, effort = 1, t_max = timetorun) #Change t_max to determine how many years the model runs for
  # 
  # env$params <-sim@params
  # 
  
          ## what kind of data and output do we have?
          if (data_type=="SSB") {
          output <-getSSB(sim)[timetorun,]   #could change to getBiomass if using survey, also check units.
          }
  
          if (data_type=="catch") {
         output <-getYield(sim)[timetorun,]/1e6 #### CHECK UNITS !! grams per year? the data are in tonnes per year so converting to tonnes.
          }
  
  pred <- log(output)
  dat  <- log(dat)

  # sum of squared errors, here on log-scale of predictions and data (could change this or use other error or likelihood options)
   discrep <- pred - dat

   discrep <- (sum(discrep^2,na.rm=T))
  
  # can use a strong penalty on the error to ensure we reach a minimum of 10% of the data (biomass or catch) for each species
  # if(any(pred < 0.1*dat)) discrep <- discrep + 1e10
  
    return(discrep)

   }

## test it


# err<-getError(vary,params)
# #err<-getError(vary,params,dat=rep(100,12),data_type="biomass")
# #test
# err
```

Now, carry out the optimisation. There are several optimastion methods to choose from - we need to select the most robust one to share here. The R package optimx seems to be the most robsut general R package and has replaced optim. Often this requires repateing the proceure several times. Further down we show how this computation can been done in paralllel.

```{r}
# this time carry out optimisation, using optim(), with catches - already saved
# vals<-optim(par=vary,getError,params=params,method ="L-BFGS-B",lower=c(rep(3,12)),upper= c(rep(20,12)))
# save params - may want to repeat this setp depending on diagnostics, after changing some parameters
# saveRDS(params,"optimx_param.RDS")
# saveRDS(sim,"optimx_sim.RDS")


# # above seems to give very different estimates every time (not sure if issues is how I am using the state, projectToSetady, or optim)
# # could also try derivative free method, for non-smaooth surfaces:
# # install.packages("dfoptim")
# library(dfoptim)

# vals<-nmkb(par=vary,fn=getError,params=steadyparams,lower=c(rep(1,12)),upper= c(rep(20,12)),tol=0.1)
# 

# comparsion of optimistaion methods, non-gradient method
library(optimx)


vals<-optimx(par=vary,getError,params=params,lower=c(rep(3,nspp)),upper= c(rep(20,nspp)), method = c("L-BFGS-B"),
control = list(trace = 6, follow.on=TRUE, maximize=FALSE))

# save vals - may want to repeat this setp depending on diagnostics, after changing some parameters
saveRDS(vals,"optimx_vals.RDS")

vals2<-optimx(par=as.numeric(vals[1:nspp]),getError,params=params,lower=c(rep(3,nspp)),upper= c(rep(20,nspp)), method = c("L-BFGS-B"),
control = list(trace = 6, follow.on=TRUE, maximize=FALSE))

#vals2<-optim(par=as.numeric(vals[1:nspp]),getError,params=params,method ="L-BFGS-B",lower=c(rep(3,nspp)),upper= c(rep(20,nspp)))

# save vals2 - may want to repeat this setp depending on diagnostics, after changing some parameters
saveRDS(vals2,"optimx_vals3.RDS")

#could repeat above to improve the fit using these values as the starting point


# plug back into model
# optimx values:
params@species_params$R_max<- 10^as.numeric(vals2[1:nspp])
# set the param object
params<-setParams(params)
sim <- project(params, effort = 1, t_max = 500, dt=0.1)
plot(sim)

# and check do species still coexist without fishing?
sim_uf <- project(params, effort = 0, t_max = 500, dt=0.1)
plot(sim_uf)


# save params - may want to repeat this setp depending on diagnostics, after changing some parameters
saveRDS(params,"optimx2_param.RDS")
saveRDS(sim,"optimx2_sim.RDS")


```

#### Step 4. Check the level ofdesnity dependence. Is the physiological recruitment, \RDI, much higher than the realised recruitment, \RDD. High RDI/RDD ratio indicates strong density dependence.

```{r}
  # params<- readRDS("optimx2_param.RDS")
  # sim <- readRDS("optimx2_sim.RDS")

getRDI(params)/getRDD(params)

# seems like there is little density dependence 

# # if needed change erepro & plug back into model
 # params@species_params$erepro[] <-1e-3
 # params <- setParams(params)
 # sim <- project(params, effort = 1, t_max = 500, dt=0.1)
 # plot(sim)

```


#### Step 5. Verify the model after the above step by comparing the model with data. 

Eg. species biomass or abundance distrubtions, feeding level, naturality mortality, growth, vulnerablity to fishing (fmsy) and catch, diet composition... Many handy functions for plotting these are available here: https://sizespectrum.org/mizer/reference/index.html


```{r}

plotPredObsYield <-function(sim,dat=sparams$yield){
## check obs vs. predicted yield
#sim<-newsim
pred_yield <-melt(getYield(sim)[100,]/1e6)
pred_yield$obs <- dat
pred_yield$species <-row.names(pred_yield)

p <- ggplot() + # plot predicted and observed yields
        geom_point(data = pred_yield, 
            aes(x = log10(value), y = log10(obs), color = species)) +
   # plot optimal fit line
        geom_abline(color = "black", slope = 1, intercept = 0) + 
  xlab("log10 Predicted Yield") + 
  ylab("log10 Observed Yield") #+ 
 # scale_fill_manual(values = wes_palette(12, "Zissou")) 
return(p)
} 

plotPredObsYield(sim) 
# check other plots

plotlyGrowthCurves(sim,percentage = T)  # check these
plotlyFeedingLevel(sim)  # low feeding levels at large sizes for some groups
plotGrowthCurves(sim,"Trichiurus_lepturus") # check each species - not reacching maturation sizes - think about the params that  control feeding, though looking at next plot they seem to accumulate in a middle range size class

plotlySpectra(sim,power=1,total = T)
plotDiet(params,"Trichiurus_lepturus") ## perhaps feeding too much on background resource, though can see a shift at larger sizes, possible low on food when it shifts to other fish?? although depends whethe your consider it ot be "other food" too, such as benthos (as in North Sea model.


# What would happen if changed erepro to increase denisty dependence?
# What happens if only vary life history params - same issues? e.g. is it the life history traits?
# same idea - setting theat to be all 1s - to help diagnose


```


#### Re-do Optimisation 

This time try again with new kappa value, and optimise on multiple cores using optimParallel.

```{r}
# read in a set of optimized params
# params<- readRDS("optim_param.RDS")
# sim <- readRDS("optim_sim.RDS")


#optim_params@resource_params$kappa <-10^5.6
#params<-setParams(steadyparams)

library("parallel")
library("optimParallel")

#sim <- newsim
vary <- log10(sim@params@species_params$R_max)
params@resource_params$kappa<-316227766017
params<-setParams(params)


cl <- parallel::makeCluster(5, setup_timeout = 0.5)
setDefaultCluster(cl = cl)
clusterExport(cl, as.list(ls()))
clusterEvalQ(cl, {
  library(mizerExperimental)
  library(optimParallel)
})

optim_result <-optimParallel(par=vary,getError,params=params,method   ="L-BFGS-B",lower=c(rep(3,12)),upper= c(rep(15,12)),
                            parallel=list(loginfo=TRUE, forward=TRUE))

stopCluster(cl)

saveRDS(optim_result,"optim_para_result.RDS")


# optim values:
params@species_params$R_max <- 10^optim_result$par 


# set the param object 
newparams<-setParams(params)
newsim <- project(params, effort = 1, t_max = 100, dt=0.1,initial_n = sim@n[100,,],initial_n_pp = sim@n_pp[100,])
plot(newsim)

plotlySpectra(newsim)
plotPredObsYield(newsim) 

## Then go back to Verification step to check plots and diagnostics

# then redo optimParallel

```


#### Step 6. The final verification step is to force the model with time-varying fishing mortality to assess whether changes in time series in biomassess and catches capture observed trends. The model will not capture all of the fluctuations from environmental processes ( unless some of these are included), but should match the magnitude and general trend in the data. We explore this in Example # 2 - Changes through time.


