---
title: "How to Mizer - how to calibrate/refine a mizer model to time-averaged catch data"
author: "Julia L. Blanchard, Ken H. Andersen, Gustav Delius, Romain Forestier"
date: "`r Sys.Date()`"
place: Hobart, Australia
output:
  html_document: default
  # pdf_document: default
always_allow_html: true
#runtime: shiny
---


```{r set up, message = F, include=FALSE}
source("../utility.R")
# loading HTM0 data
# model
sim_guessed <- readRDS("../HTM1/HTM1_sim.rds")
params_guessed <- sim_guessed@params
# database
# fisheries mortality F
fMat <- read.csv("../data/fmat.csv")
# fMatWeighted <- read.csv("data/fmatWeighted.csv") # Sandeel and Cod have multiple data base so average their F and weighting by SSB
fMatWeighted <- readRDS("../data/FmatWeightedInterpolated.rds") # to get Gurnard data
# read in time-averaged  catches  
catchAvg <-read.csv("../data/time-averaged-catches.csv") # only that one is used at the moment | catches are estimated from fMatW
# ssb
ssbAvg <- read.csv("../data/time-averaged-SSB.csv")
```


# Intermediate level tutorial - how to calibrate a Mizer model

In this tutorial you will learn 

- How to refine key parameters using optimization so the model is in the right ballpark as the data

- Simple calibration protocol - shown sequentially for `R_max`, `erepro` and `kappa`

- Introduce rShiny for model exploration


In this tutorial, we are going to follow the same path as HTM1 and keep using the North Sea data set as an example.

We finished HTM1 with a manually parametrised model with species coexisting, now we want our model to reflect reality. To do so we are going to compare the model output to data by matching catch, biomass and/or growth rate depending on what is available.

In the plot below we compare the catch output of the model and the averaged catch we extracted from the ICES database in the previous tutorial using the function `plotPredObsYield`. This is going to be our main comparison used to calibrate the model towards more real life behavior.


```{r step 0 - observed/predicted catches}
plotPredObsYield(sim_guessed,catchAvg$Catch_1419_tonnes) 
```

At the moment, the observed yield are higher than the predicted yield for all species. We want them to be equal (on the diagonal).



## Step 1. Calibrate the maximum recruitment

In this section you will:

- use a package that will calibrate `R_max` per species

`R_max` affects the relative biomass of each species (and, combined with the fishing parameters, the catches). We are going to change with the aim of minimising the error between observed and estimated catches or biomasses. Both `erepro` and $\kappa$ affect the relative biomass of each species and can be used (as in Blanchard et al. 2014 & Spence et al. 2016) but for now we are going to keep them at default value and focus on `R_max` only.

First let's set up a function running the model and outputing the difference between predicted catches (`getYield()`) and actual catches (`catchAvg`). `getError()` outputs the sum of squared errors between the two.

```{r step step 1 - getError | function, include=F}
## the following getError function combines the steps of the optimisastion above - this time with the multispecies model and output the predicted size spectrum
## update below with project_steady and saving the state from each iteration
#RF the function takes a bunch of RMax and compare the theoretical catches versus data
getError <- function(vary,params,dat,env=state,data_type="catch", tol = 0.1,timetorun=10) {
  
  #env$params@species_params$R_max[]<-10^vary[1:12]
  params@species_params$R_max[]<-10^vary[1:12]
  
  params <- setParams(params)
  # run to steady state and update params
  # env$params<- projectToSteady(env$params, distance_func = distanceSSLogN,
  #                 tol = tol, t_max = 200,return_sim = F)
  params<- projectToSteady(params, distance_func = distanceSSLogN,
                   tol = tol, t_max = 200,return_sim = F)
 
  # create sim object 
   
  sim <- project(params, effort = 1, t_max = timetorun) #Change t_max to determine how many years the model runs for
  
  # 
  # sim <- project(env$params, effort = 1, t_max = timetorun) #Change t_max to determine how many years the model runs for
  # 
  # env$params <-sim@params
  # 
  
          ## what kind of data and output do we have?
          if (data_type=="SSB") {
          output <-getSSB(sim)[timetorun,]   #could change to getBiomass if using survey, also check units.
          }
  
          if (data_type=="catch") {
         output <-getYield(sim)[timetorun,]/1e6 
         #' using n . w . dw so g per year per volume (i.e. North Sea since kappa is set up this way). 
         #'The data are in tonnes per year so converting to tonnes.
          }
  
  pred <- log(output)
  dat  <- log(dat)
  # sum of squared errors, here on log-scale of predictions and data (could change this or use other error or likelihood options)
   discrep <- pred - dat
   discrep <- (sum(discrep^2))
  
  # can use a strong penalty on the error to ensure we reach a minimum of 10% of the data (biomass or catch) for each species
  # if(any(pred < 0.1*dat)) discrep <- discrep + 1e10
  
    return(discrep)
   }
```

```{r step 1 - getError | result}
# we need 12 Rmaxs, log10 scale
vary <- log10(params_guessed@species_params$R_max)
## set up the environment to keep the current state of the simulations 
state <- new.env(parent = emptyenv())
state$params <-  params_guessed
## test it
getError(vary = vary, params = params_guessed, dat = catchAvg$Catch_1419_tonnes)

```


Now, carry out the optimisation. There are several optimisation methods to choose from - we need to select the most robust one to share here. The R package `optimParallel` seems to be the most robust general R package and has replaced optim. Often this requires repeating the procedure several times but the advantage of using parallel run is the speed compared to packages such as optimx.

This might take AWHILE. The output is saved as `optimParallel_Rmax.RDS` if you wish to skip this block.

```{r step 1 - optimisation, message = F, eval=F}
# create set of params for the optimisation process
params_optim <- params_guessed
vary <-  log10(params_optim@species_params$R_max) # variable to explore
params_optim<-setParams(params_optim)
# set up workers
noCores <- detectCores() - 1 # keep some spare core
cl <- makeCluster(noCores, setup_timeout = 0.5)
setDefaultCluster(cl = cl)
clusterExport(cl, as.list(ls()))
clusterEvalQ(cl, {
  library(mizerExperimental)
  library(optimParallel)
})

# tic()
optim_result <-optimParallel(par=vary,getError,params=params_optim, dat = catchAvg$Catch_1419_tonnes, method   ="L-BFGS-B",lower=c(rep(3,12)),upper= c(rep(15,12)),
                            parallel=list(loginfo=TRUE, forward=TRUE))
stopCluster(cl)
# toc() # 80'' using 47 cores
saveRDS(optim_result,"optimParallel_Rmax1.RDS")
```


```{r step 1 - results, message=FALSE,warning=FALSE}
# if previous block was not evaluated
params_optim <- params_guessed
optim_result <- readRDS("optimParallel_Rmax1.RDS")
# optim values:
params_optim@species_params$R_max <- 10^optim_result$par 
# set the param object 
params_optim <-setParams(params_optim)
sim_optim <- project(params_optim, effort = 1, t_max = 100, dt=0.1,initial_n = sim_guessed@n[100,,],initial_n_pp = sim_guessed@n_pp[100,])
saveRDS(sim_optim,"sim_optim1.RDS")
plotSummary(sim_optim, short = T)
plotPredObsYield(sim_optim,catchAvg$Catch_1419_tonnes) 
```

Calibrating for best observed/predicted yield makes one species (Sprat) show signs of collapse. We need to look at other parameters to get the community to coexist again.



## Step 2. Calibrate the recruitment with erepro.

In this section you will:

- Look at effect of `erepro` on the reproductive outputs

- Check what impact `erepro` has on the $F_{msy}$


`erepro` represents the energy conversion efficiency between energy allocated to reproduction and the actual spawn. Lowering `erepro` biologically means higher egg mortality rate or wasteful energy invested into gonads. For example, `erepro` is currently set to $0.01$ meaning, that for every one $g$ of mass allocated to reproduction, $0.01g$ will be used as spawn recruitment, the rest is lost. This coefficient is applied before any density-dependence requirement. Let's use Rshiny to see how varying `erepro` influences the ecosystem (to play around).


> Comment: shiny disabled for pdf/html, use in Rmd only


```{r step 2 r-shiny, eval = F, cache = F, include = F}
# Optional
sim_optim <- readRDS("sim_optim1.RDS")

# adjust Rmax and/or reproductive efficiency to examine whether improved steady state is achievable that way
library(shiny) # no need if runtime = shiny is in the YAML
# runApp("shiny-equilibrium")
# is there a way to save the final chosen values?
params_shiny <- sim_optim@params
shinyApp(
  ui=fluidPage(
  
  # Application title
  titlePanel("North Sea Model Example"),
  
  fluidRow(
    column(4, wellPanel(
       #sliderInput("kappa", "log10 Resource Carrying Capacity:", min = 8, max = 12, value = log10(params_shiny@resource_params$kappa),
       #            step = 0.1),
    #   sliderInput("Rmax", "log10 Maximum Recruitment:", min = 1, max = 12, value = 12,
    #              step = 0.1),
       sliderInput("erepro", "log10 Reproductive Efficiency:", min = -8, max = 1, value = log10(params_shiny@species_params$erepro[1]),
                   step = 0.1)
          )),
    column(6,
           plotOutput("distPlot", width = 600, height = 600)
    ))
     
  
    
  ),
  
  server = function(input, output) {
   
  output$distPlot <- renderPlot({
    # set up params using values given, need check and change parameter values so units work in days units 
    params_shiny@species_params$erepro <- rep(10^input$erepro,12)
   # params@species_params$Rmax <- rep(10^input$Rmax,12)
    params_shiny <- setParams(params_shiny)#,kappa=10^input$kappa)
    # run without fishing
    sim_shiny <- project(params_shiny, effort = 1, t_max = 50)
    plot(sim_shiny)
     })
},
  options = list(height = 500)
)
```


A good indicator of realistic `erepro` value is the shape of the fishing mortality rate versus yield curve. Such graph allows to determine the fisheries maximum sustainbale yield when harvesting a species and is supposed to have a bell shape. Too high fishing mortality depletes the stock and reduce the yield whereas too low fishing mortality simply means low yield. The $F_{msy}$ sits in the middle, at the top of the bell. If a species is too reproduction efficient and can replenish its biomass even under high fisheries mortality, its `erepro` is too high. Conversely, if a low fisheries mortality depletes the stock right away, the species' `erepro` is too low.

Let's take a look at the current $F_{msy}$ per species. 

```{r step 2 - Fmsy | data crunching, eval=F}
# Not evaluated because long (especially at high effortRes)
plot_dat <- plotFmsy(params_optim, returnData = T, effortRes = 50)
saveRDS(plot_dat, "Fmsy1.rds")
```


```{r step 2 - Fmsy | plots, warning=F}
plot_dat <- readRDS("Fmsy1.rds")
ggplot(plot_dat) +
 geom_line(aes(x = effort , y = yield, color = species))+
  facet_wrap(species~., scales = "free") +
  scale_x_continuous(limits= c(0,1.5),name = "fishing mortality rate")+#, limits = c(1e10,NA))+
  scale_y_continuous(trans = "log10") +
  scale_color_manual(name = "Species", values = sim_optim@params@linecolour) +
    theme(legend.position = "none", legend.key = element_rect(fill = "white"),
           panel.background = element_blank(), panel.grid.minor = element_line(color = "gray"),
          strip.background = element_blank())
```

Decreasing `erepro` is going to move the $F_{msy}$ towards lower effort. Until now we had the same `erepro` for all species but we can input species-specific values to calibrate our recruitment. To do so, let's use another shiny app again. The next one allows you to change `erepro` one at a time and see the effect on the species' $F_{msy}$

> Comment: shiny disabled for pdf/html, use in Rmd only
> Note: the shiny app only calculate the $F_{msy}$ of one species at a time (for speed reasons) and the final results may not be exactly the same as calculating the $F_{msy}$ for every species together (as shown below)

```{r step 2 - shiny Fmsy, eval = F, cache = F, include = F}
# Optional
sim_optim <- readRDS("sim_optim1.RDS")
# adjust Rmax and/or reproductive efficiency to examine whether improved steady state is achievable that way
library(shiny) # no need if runtime = shiny is in the YAML
# runApp("shiny-equilibrium")
# is there a way to save the final chosen values?
params_shiny <- sim_optim@params
FmsyDat <- readRDS("Fmsy1.rds")
shinyApp(
  ui=fluidPage(
  
  # Application title
  titlePanel("North Sea Fmsy"),
  
  fluidRow(
    column(4, wellPanel(
       # sliderInput("kappa", "log10 Resource Carrying Capacity:", min = 8, max = 12, value = log10(params_optim@resource_params$kappa),
       #             step = 0.1),
    #   sliderInput("Rmax", "log10 Maximum Recruitment:", min = 1, max = 12, value = 12,
    #              step = 0.1),
       sliderInput("erepro", "log10 Reproductive Efficiency:", min = -8, max = 1, value = log10(sim_optim@params@species_params$erepro[1]),
                   step = 0.1),
sliderTextInput(
    inputId = "species",
    label = "Species name",
    choices = params_shiny@species_params$species,
    selected = params_shiny@species_params$species[1],
    grid = T
  ),
          )),
    
    column(6,
           plotOutput("distPlot", width = 600, height = 600)
    ))
     
  
    
  ),
  
  server = function(input, output) {
   
  output$distPlot <- renderPlot({
    # set up params using values given, need check and change parameter values so units work in days units 
    params_shiny@species_params$erepro[which(params_shiny@species_params$species == input$species)] <-10^(input$erepro)
    
    plotFmsy(params_shiny, speciesData = list(input$species,FmsyDat), effortRes = 10)
     })
},
  options = list(height = 500)
)
# add species slider to display jsut one species at a time
```


Let's see what our system looks like with species-specific `erepro`

```{r step 2 - Fmsy with tweaked erepro, warning=F}
sim_optim <- readRDS("sim_optim1.RDS")
params_optim <- sim_optim@params # redundancy if previous blocks were not run or something
params_optim@species_params$erepro <- 10^(c(-1.7,-3.2,-4,-4,-3,-3.5,-3.5,-4.1,-2.7,-3,-2.5,-1.7))
params_optim2 <- setParams(params_optim)
sim_optim2 <- project(params_optim2, effort = 1, t_max = 100)
saveRDS(sim_optim2, file = "sim_optim2.RDS")
```


```{r step 2 - Fmsy plot 2, warning=F, eval = F}
plot_dat <- plotFmsy(params_optim2, effortRes = 50, returnData = T)
saveRDS(plot_dat, "Fmsy2.rds")
```


```{r step 2 - Fmsy with tweaked erepro summary, warning=F}

plotSummary(sim_optim2, short = T)

plot_dat <- readRDS("Fmsy2.rds")
ggplot(filter(plot_dat)) +
 geom_line(aes(x = effort , y = yield, color = species))+
  facet_wrap(species~., scales = "free") +
  scale_x_continuous(limits= c(0,1.5),name = "fishing mortality rate")+#, limits = c(1e10,NA))+
  scale_y_continuous(trans = "log10") +
  scale_color_manual(name = "Species", values = sim_optim@params@linecolour) +
    theme(legend.position = "none", legend.key = element_rect(fill = "white"),
           panel.background = element_blank(), panel.grid.minor = element_line(color = "gray"),
          strip.background = element_blank())

```

Some trade-offs here, relatively high `erepro` allows Sprat to coexist with the other species but the $F_{msy}$ plot doesn't have the characteristic bell shape. Here Sprat is too efficient and survive even at high fisheries mortality. Note that the curve could start to go down at higher fisheries mortality.

Before spending more tome hand tweaking these species, we are going to run the `R_max` calibration again to see if we improved the yield match (as in step 1).

Itering the calibration process multiple times can help find better `R_max` values as the first run of `optimParallel` may not find the best combination of `R_max` in the first go. Below it an example of `optimParallel` being looped five times with 

```{r step 2 - looping rmax calibration crunch, message = F, eval=F}

# looping calibration to get better result
sim_loop <- readRDS(file = "sim_optim2.RDS")
params_loop <- sim_loop@params
for(i in 1:5)
{
# saving the last time step of sim_optim2 in the param object
params_loop@initial_n <- sim_loop@n[dim(sim_loop@n)[1],,]
params_loop@initial_n_pp <- sim_loop@n_pp[dim(sim_loop@n_pp)[1],]
params_calibration <- params_loop
vary <-  log10(params_calibration@species_params$R_max)
params_calibration<-setParams(params_calibration)
noCores <- detectCores() - 1 # keep a spare core
cl <- makeCluster(noCores, setup_timeout = 0.5)
setDefaultCluster(cl = cl)
clusterExport(cl, as.list(ls()))
clusterEvalQ(cl, {
  library(mizerExperimental)
  library(optimParallel)
})
tic()
optim_loop <-optimParallel(par=vary,getError,params=params_calibration, dat = catchAvg$Catch_1419_tonnes, 
                             method   ="L-BFGS-B",lower=c(rep(3,12)),upper= c(rep(15,12)),
                            parallel=list(loginfo=TRUE, forward=TRUE))
stopCluster(cl)
toc() # 80'' using 47 cores



# optim values:
params_loop@species_params$R_max <- 10^optim_loop$par 
# set the param object 
params_loop <-setParams(params_loop)
sim_loop <- project(params_loop, effort = 1, t_max = 100, dt=0.1, initial_n = params_loop@initial_n ,
                      initial_n_pp = params_loop@initial_n_pp)
}
saveRDS(optim_loop,"optimParallel_Rmax2.RDS")

```

```{r saving sim, eval = F}
# running sim a bit longer before saving it
sim_loop <- project(params_loop, effort = 1, t_max = 300)
params_loop@initial_n <- sim_loop@n[dim(sim_loop@n)[1],,]
params_loop@initial_n_pp <- sim_loop@n_pp[dim(sim_loop@n_pp)[1],]
sim_loop <- project(params_loop, effort = 1)
saveRDS(sim_loop,"sim_optim3.RDS")
```


```{r sim loop diag, eval = T}

sim_loop <- readRDS("sim_optim3.RDS")
plotSummary(sim_loop, short = T)
plotPredObsYield(sim_loop,catchAvg$Catch_1419_tonnes) 



```


The different species look more spread around the x = y line but the scale is also three times smaller than the previous plot. Here, only Sandeel is an outlier.

## Step 3. Calibrating the growth

In this section you will:

- look at the growth curves of each species

- tweak the $\kappa$ parameter to adjust the growth curves


```{r step 3 - growth curves prior}
plotGrowthCurves2(sim_loop, species_panel = T)
# Ken's comment: plotGrowthCurves. Legend should be “model”, “Observed”. Need to discuss about that since we need to change the plot in Mizer then
```

Most species have a growth curve similar to the expected von Bertalanffy growth curve, apart from Spart and Sandeel which have much slower growth than expected. $\kappa$, which is the carrying capacity of the background spectrum will affect the food availability and therefore the growth rate (more food means faster growth). However, changing $\kappa$ will affect all growth curves. If only a few species are off, we need to change the $\gamma$ parameter (search volume) per species.



```{r step 3 - feeding level}
plotFeedingLevel(sim_loop, include_critical = F)
```

The feeding level here shows that the species reaching the highest size classes have a feeding level close to one, meaning that they feed to satiation. Let's look at their diet to check what makes them to full.

```{r step 3 - diets, warning=F}
# plotDiet2(sim_optim2, "Haddock")
# plotDiet2(sim_optim2, "Plaice")
# plotDiet2(sim_optim2, "Saithe")
plotDiet2(sim_loop)
```

The diets look great as the fish feed on each other and do not rely too much on the background spectrum. We can still reduce the carrying capacity of the background spectrum even more, which could lower the feeding level of the species. Let's do this using the Rshiny app


RF: Rshiny section disabled for pdf

```{r step 3 r-shiny, eval = F, cache = F, include = F}
# Optional
# adjust Rmax and/or reproductive efficiency to examine whether improved steady state is achievable that way
library(shiny) # no need if runtime = shiny is in the YAML
# runApp("shiny-equilibrium")
# is there a way to save the final chosen values?
params_shiny <- sim_optim2@params
shinyApp(
  ui=fluidPage(
  
  # Application title
  titlePanel("North Sea Model Example"),
  
  fluidRow(
    column(4, wellPanel(
       sliderInput("kappa", "log10 Resource Carrying Capacity:", min = 8, max = 12, value = log10(params_shiny@resource_params$kappa),
                   step = 0.1),
    #   sliderInput("Rmax", "log10 Maximum Recruitment:", min = 1, max = 12, value = 12,
    #              step = 0.1),
       # sliderInput("erepro", "log10 Reproductive Efficiency:", min = -8, max = 1, value = -2,
       #             step = 0.1)
          )),
    column(6,
           plotOutput("distPlot", width = 600, height = 600)
    ))
     
  
    
  ),
  
  server = function(input, output) {
   
  output$distPlot <- renderPlot({
    # set up params using values given, need check and change parameter values so units work in days units 
    # params_shiny@species_params$erepro <- rep(10^input$erepro,12)
   # params@species_params$Rmax <- rep(10^input$Rmax,12)
    params_shiny <- setParams(params_shiny,kappa=10^input$kappa)
    # run without fishing
    sim_shiny <- project(params_shiny, effort = 1, t_max = 100)
    plot(sim_shiny)
     })
},
  options = list(height = 500)
)
```

Decreasing slightly $\kappa$ drives Sprat to extinction probably as it becomes the next choice of food when we reduce the size of the background spectrum (as we saw on the diet plots). So how can we reduce the feeding level of the largest predators while keeping Sprat alive?

```{r step 3 - kappa tweak, include = F, eval = F}
# updating param parameters
params_optim2@resource_params$kappa<- 10^(11.5)
params_optim2 <-setParams(params_optim2)
sim_optim2 <- project(params_optim2, effort = 1, t_max = 100, dt=0.1)#,initial_n = sim_optim@n[100,,],initial_n_pp = sim_optim@n_pp[100,])
# saveRDS(sim_optim,"optim_para_sim2.RDS")
plotGrowthCurves(sim_optim2, species_panel = T)
# keep last time step of sim_optim saved to not have to get Sprat coming back from the deads every time
params_optim2@initial_n <- sim_optim2@n[dim(sim_optim2@n)[1],,]
params_optim2@initial_n_pp <- sim_optim2@n_pp[dim(sim_optim2@n_pp)[1],]
```