---
title: Why use Mizer in the first place?
author: RF
output: html_document
---


Mizer is a tool that can simulate a dynamic size spectrum in an marine ecosystem, subject to fishing pressure. 
Multiple fishing gears can be defined and fishing mortality can change through time allowing to explore a range of fishing patterns / exploitation strategies and their effect on the marine ecosystem.

Below are some examples of what one can do with a calibrated Mizer model.

```{r set-up, echo = F,message=F, warning=F}
source("../utility.R")
sim <- readRDS("../HTM2/sim_optim3.RDS")

```

### The Nort Sea ecosystem

```{r plotSummary, echo = F, warning=F}
plotSummary(sim)
```




### Varying the fisheries effort

The effort directly scales the fishing mortality rate.

```{r gear set-up, echo = F}
  # make one gear per species so we can vary the effort per species
  gear <- gear_params(sim@params)
  gear$gear <- sim@params@species_params$species
    # prepping species_params to handle double sigmoid and 
  gear$l50_right <- sim@params@species_params$l50 + 2*(sim@params@species_params$l50 - sim@params@species_params$l25)
  gear$l25_right <- gear$l50_right + sim@params@species_params$l50 - sim@params@species_params$l25
  gear$sigmoidal_weight <- .5*sim@params@species_params$w_inf
  gear$sigmoidal_sigma <- 10
  gear_params(sim@params) <- gear
  
```


```{r varying effort, eval = T, echo = F, warning=F}


plot_dat <- NULL
for(iScenario in 1:4)
{
  switch(iScenario,
         "1" = {ieffort = c(1,1,1,1,1,1,1,1,2,2,2,2)
         scenario_name = "overfishing big"},
         "2" = {ieffort = c(2,2,2,2,1,1,1,1,1,1,1,1)
         scenario_name = "overfishing_small"},
         "3" = {ieffort = c(0,0,0,0,0,0,0,0,1,1,1,1)
         scenario_name = "big_only"},
         "4" = {ieffort = c(2,2,2,2,0,0,0,0,1,1,1,1)
         scenario_name = "medium_safe"},
         {}
  )
  
  
  sim <- project(sim@params, effort = ieffort, t_max = 50)
  fDat <- plotFMort(sim, return_data = T)
fDat$scenario <- scenario_name
fDat$type <- "Fishing mortality"

sDat <- plotSpectra(sim, return_data = T)
sDat <- sDat[[1]]
sDat$scenario <- scenario_name
sDat$type <- "Size Spectrum"
sDat$value <- log10(sDat$value)

plot_dat <- rbind(plot_dat,fDat,sDat)
  
}

ggplot(plot_dat) +
  geom_line(aes(x = w, y = value, color = Species ))+
  scale_x_continuous(trans = "log10", limits = c(.1,NA), name = "Size") +
  scale_color_manual(values = sim@params@linecolour) +
  scale_y_continuous(name = "Biomass density / fishing mortality rate") +
  facet_grid(type~scenario, scales = "free_y") +
  theme(strip.background = element_blank(),
        panel.background = element_blank(),
        panel.grid.minor = element_line(color = "gray"),
        panel.border = element_rect(colour = "gray", fill=NA, size=.5),
        legend.position = "bottom", legend.key = element_rect(fill = "white"))


```


> Rshiny app for effort (only available on .rmd)

```{r step 2 - shiny Fmsy, eval = F, cache = F, include = F}
# Optional

# adjust Rmax and/or reproductive efficiency to examine whether improved steady state is achievable that way
library(shiny) # no need if runtime = shiny is in the YAML
# runApp("shiny-equilibrium")
# is there a way to save the final chosen values?
params_shiny <- sim@params
effort_vec <- rep(1,length(sim@params@species_params$species))
shinyApp(
  ui=fluidPage(
  
  # Application title
  titlePanel("North Sea fisheries"),
  
  fluidRow(
    column(4, wellPanel(
       # sliderInput("kappa", "log10 Resource Carrying Capacity:", min = 8, max = 12, value = log10(params_optim@resource_params$kappa),
       #             step = 0.1),
    #   sliderInput("Rmax", "log10 Maximum Recruitment:", min = 1, max = 12, value = 12,
    #              step = 0.1),
       sliderInput("effort", "fisheries effort:", min = 0, max = 5, value = 1, step = 0.1),
sliderTextInput(
    inputId = "species",
    label = "Species name",
    choices = params_shiny@species_params$species,
    selected = params_shiny@species_params$species[1],
    grid = T
  ),
#textOutput("selected_var")
          )),
    
    column(6,
           plotOutput("distPlot", width = 600, height = 600)
    ))
     
  
    
  ),
  
  server = function(input, output) {
    

    
    
   
  output$distPlot <- renderPlot({
    # set up params using values given, need check and change parameter values so units work in days units 
    
     effort_vec[which(params_shiny@species_params$species == input$species )] <- input$effort
    sim_shiny <- project(params_shiny, t_max = 25, effort = effort_vec)

    plot_dat1 <- plotSpectra(sim_shiny, power = 1, wlim = c(.1,NA), return_data = TRUE, total = TRUE)
  p1 <- ggplot(plot_dat1[[1]]) +
    geom_line(aes(x = w, y = value, colour = Species, group = Species)) +
    scale_x_continuous(limits = c(.1,NA), trans = "log10", name = "Individual size [g]")+#, breaks = log_breaks()) +
    scale_y_continuous(name = "Biomass density" ,trans = "log10", breaks = log_breaks()) +
    # scale_y_continuous(name = expression(paste("Biomass density (ind.", m^{-3},")", sep="")) ,trans = "log10", breaks = log_breaks()) +
    scale_colour_manual(values = sim@params@linecolour) +
    scale_linetype_manual(values = sim@params@linetype) +
    theme(axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          # text = element_text(size=font_size),
          panel.background = element_blank(),
          panel.grid.minor = element_line(color = "gray"),
          panel.border = element_rect(colour = "gray", fill=NA, size=.5),
          legend.position = "none", legend.key = element_rect(fill = "white"))
  
  plot_dat2 <- plotFMort(sim_shiny, return_data = T)
      p2 <-   ggplot(plot_dat2) +
      geom_line(aes(x = w, y = value, colour = Species)) +
      scale_x_continuous(name = "Individual size [g]", trans = "log10", limits = c(.1,NA)) +
      scale_y_continuous(name = "Fisheries mortality [1/year]", limits = c(.01, max(plot_dat2$value))) +
      scale_colour_manual(values = sim@params@linecolour) +
      theme(panel.background = element_blank(),
            panel.grid.minor = element_line(color = "gray"),
            panel.border = element_rect(colour = "gray", fill=NA, size=.5),
            # text = element_text(size=font_size),
            legend.position = "bottom", legend.key = element_rect(fill = "white")) +
           guides(color = guide_legend(nrow=2))
  
    
     # plot_grid(p1,p2, mylegend,rel_heights = c(5,5,1), nrow = 3,align = "v")#, axis = "l")
           plot_grid(p1,p2, nrow=2,align = "v")#, axis = "l")

     })
  #output$selected_var <- renderText({ as.character(effort_vec)})
},
  options = list(height = 500)
)

# effort_vec is reset everytime
# need to make as many sliders as species I guess
# add species slider to display jsut one species at a time
```


### Varying the selectivity function

The selectivity function determines which size class are targeted by the fisheries. Some selectivity function are available in Mizer such as: `knife_edge()`, `sigmoid_length()`, `double_sigmoid_length()` and `sigmoid_weight()`. The user can input its own selectivity function.
 
```{r fishing gear, warning=F, echo = F}
gear <- gear_params(sim@params)
plot_dat <- NULL
gear_vec <-  c("knife_edge", "sigmoid_length", "double_sigmoid_length", "sigmoid_weight")
for(iGear in gear_vec)
{
gear$sel_func <- iGear
gear_params(sim@params) <- gear

sim <- project(sim@params,t_max = 50,effort = 2)

fDat <- plotFMort(sim, return_data = T)
fDat$gear <- iGear
fDat$type <- "Fishing mortality"

sDat <- plotSpectra(sim, return_data = T)
sDat <- sDat[[1]]
sDat$gear <- iGear
sDat$type <- "Size Spectrum"
sDat$value <- log10(sDat$value)

plot_dat <- rbind(plot_dat,fDat,sDat)
}

ggplot(plot_dat) +
  geom_line(aes(x = w, y = value, color = Species ))+
  scale_x_continuous(trans = "log10", limits = c(.1,NA), name = "Size") +
  scale_color_manual(values = sim@params@linecolour) +
  scale_y_continuous(name = "Biomass density / fishing mortality rate") +
  facet_grid(type~gear, scales = "free_y") +
  theme(strip.background = element_blank(),
        panel.background = element_blank(),
        panel.grid.minor = element_line(color = "gray"),
        panel.border = element_rect(colour = "gray", fill=NA, size=.5),
        legend.position = "bottom", legend.key = element_rect(fill = "white"))


```
 
 